---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
attach(brfss2013)
```



* * *

## Part 1: Data
Random sampling?

"The survey is conducted using Random Digit Dialing (RDD) techniques on both landlines and cell phones." Source: https://www.cdc.gov/brfss/about/brfss_faq.htm
Sampling appears to be random, therefore, results can be generalized to the population of the US.However, there might be some selection bias introduced due to the survey being telephone-based. A proportion of the population of interest might not own telephones; landlines or cellular.

Random assignment?

Since this is a cross-sectional study, it is observational, not interventional, So random assignment is not possible. Therefore, we cannot establish causality, only associations.

* * *

## Part 2: Research questions

**Research quesion 1:**

My parents are both doctors and my mother sleeps for much longer than my father!
Does your biological sex affect how long you sleep?
Which sex sleeps longer ? (grouping both sexes 
by employment status to reduce the effect of employment as a confounder).

**Research quesion 2:**

Does the prevalence of high blood cholesterol affect that of heart diseases or strokes?
According to the Mayo Clinic, high cholesterol is a risk factor for heart disease and strokes.Does this data confirm that?
Source: https://www.mayoclinic.org/diseases-conditions/high-blood-cholesterol/symptoms-causes/syc-20350800

**Research quesion 3:**

Does satisfaction with life increase with income?
Since I'm currently broke, I'd like to know.

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**

Difference between males and females in mean sleeping duration and average alcohol drinking per day in the last 30 days.

```{r}
brfss2013 <- brfss2013 %>%
 filter(!is.na(addepev2), !is.na(sleptim1), !is.na(avedrnk2))
```
```{r}
brfss2013 %>%
  group_by(addepev2) %>%
  summarise(mean_sleep = mean(sleptim1), mean_av_alcohol = mean(avedrnk2))
```

```{r}
ggplot(data = brfss2013, aes(x = avedrnk2, y = sleptim1)) +
  geom_point() + geom_smooth(method = "lm", colour = "red") +
  facet_grid(~ addepev2)
```

```{r}
brfss2013 <- brfss2013 %>%
  filter(!is.na(educa), !is.na(avedrnk2), !is.na(sex))
```

```{r}
ggplot(data = brfss2013, aes(x = educa, y = avedrnk2, fill = educa)) +
  geom_col() +
  labs(x = "Level of Education", y = "Average Alcohol Intake", fill ="Education") +
  ggtitle("Relationship between Education and Alcohol Consumption") +
  scale_x_discrete(labels = abbreviate) +
  facet_grid(~ sex) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
ggplot(data = brfss2013, aes(x = educa , y = avedrnk2, fill = educa)) +
  geom_col(na.rm = T) +
  labs(x = "Level of Education", y = "Average Alcohol Intake", fill ="Education") +
  ggtitle("Relationship between Education and Alcohol Consumption") +
  scale_x_discrete(labels = abbreviate) +
  scale_y_continuous(limits = c(0,2000)) +
  facet_grid(~ sex) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
```{r}
ggplot(data = brfss2013, aes(x = educa , y = avedrnk2, fill = educa)) +
  geom_col(na.rm = T) +
  labs(x = "Level of Education", y = "Average Alcohol Intake", fill ="Education") +
  ggtitle("Relationship between Education and Alcohol Consumption") +
  scale_x_discrete(labels = abbreviate) +
  expand_limits(y = c(0,80)) +
  facet_grid(~ sex) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```



```{r}
brfss2013 <- brfss2013 %>%
 filter(!is.na(X_bmi5cat), !is.na(cvdcrhd4), !is.na(toldhi2))
```

```{r}
ggplot(brfss2013, aes(x = toldhi2, fill = cvdinfr4)) +
  geom_bar(position = "fill",) +
  labs(x = "High Cholesterol", y = "Heart Attack") +
  ggtitle("Associaton between High Cholesterol and Heart Attacks") +
  scale_fill_discrete(name = "High Cholesterol") +
  facet_grid(~ X_bmi5cat) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


**Research quesion 2:**

Relationship between frequency of fruit intake and number of days with poor physical health
```{r}
brfss2013 %>%
  mutate(men = as.numeric(as.character(nummen)), women = as.numeric(as.character(numwomen))) %>%
  filter(!is.na(men), !is.na(women), !is.na(sex)) %>%
  group_by(sex) %>%
  summarise(meanmen = mean(men), meanwomen = mean(women))
ggplot(brfss2013, aes(x= men, y= women))+
  geom_point()
```

```{r}
brfss2013 <- brfss2013 %>%
  filter(!is.na(educa), !is.na(checkup1), !is.na(sex))
```
```{r}
library(ggthemes)
ggplot(data = brfss2013, aes(x = educa , y = checkup1, fill = educa)) +
  geom_count() + 
  labs(x = "Level of Education", y = "Time Since Last Checkup", fill = "Education") +
  ggtitle("El") +
  facet_grid(~ sex) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
brfss2013 <- brfss2013 %>%
  filter(!is.na(prediab1), !is.na(fruit1), !is.na(sex))
library(ggthemes)
ggplot(data = brfss2013, aes(x = prediab1 , y = fruit1, fill = prediab1)) +
  geom_col() + 
  labs(x = "Level of Education", y = "Average Alcohol Intake", fill = "Education") +
  ggtitle("Relationship between Education and Alcohol Consumption") +
  scale_fill_discrete(name = "Education") +
  facet_grid(~ sex) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
brfss2013 <- brfss2013 %>%
  filter(!is.na(toldhi2))
ggplot(data = brfss2013, aes(x = bphigh4, y = cvdstrk3)) +
  geom_col(na.rm = TRUE) +
  facet_grid(~ toldhi2)
```

```{r}
brfss2013 <- brfss2013 %>%
  filter(!is.na(sex), !is.na(fruit1), !is.na(physhlth))
brfss2013 %>%
  group_by(sex) %>%
  summarise(meanfruit = mean(fruit1), meanphysical_notgood = mean(physhlth))
```

```{r}
ggplot(data = brfss2013, aes(x = fruit1, y = physhlth)) +
  geom_point() + geom_smooth(method = "lm", colour = "red") +
  facet_grid(~ sex)
```

```{r}

```

**Research quesion 3:**

The relationship between income and satisfation with life

```{r}
brfss2013 <- brfss2013 %>%
  filter(!is.na(income2), !is.na(lsatisfy))
ggplot(brfss2013, aes(x = income2, fill = lsatisfy)) +
  geom_bar(position = "fill",) +
  labs(x = "Income Level", y = "Satisfaction with Life") +
  ggtitle("Associaton between Income and Satisfaction with life") +
  scale_fill_discrete(name = "Satisfaction with Life") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
v3 <- table(income2, lsatisfy)
plot(v3)
```

